---
title: "Stawberries"
author: Truc Minh Nguyen
date: 2024 Oct 11
format: html
editor: visual
---

# Preparing data for analysis

Acquire, explore, clean & structure, EDA

### Data cleaning and organization

["An introduction to data cleaning with R" by Edwin de Jonge and Mark van der Loo](https://cran.r-project.org/doc/contrib/de_Jonge+van_der_Loo-Introduction_to_data_cleaning_with_R.pdf)

["Problems, Methods, and Challenges in Comprehensive Data Cleansing" by Heiko Müller and Johann-Christoph Freytag](https://www.researchgate.net/profile/Heiko-Mueller/publication/228929938_Problems_methods_and_challenges_in_comprehensive_data_cleansing/links/09e415101b58541e2c000000/Problems-methods-and-challenges-in-comprehensive-data-cleansing.pdf)

## Strawberries

**Questions**

-   Where they are grown? By whom?

-   Are they really loaded with carcinogenic poisons?

-   Are they really good for your health? Bad for your health?

-   Are organic strawberries carriers of deadly diseases?

-   When I go to the market should I buy conventional or organic strawberries?

-   Do Strawberry farmers make money?

-   How do the strawberries I buy get to my market?

## The data

The data set for this assignment has been selected from:

\[[USDA_NASS_strawb_2024SEP25](https://quickstats.nass.usda.gov) The data have been stored on NASS here: [USDA_NASS_strawb_2024SEP25](https://quickstats.nass.usda.gov/results/28141CE5-7429-36C3-A385-33ACC8030763)

and has been stored on the blackboard as strawberries25_v3.csv.

## read and explore the data

Set-up

```{r}
#| label: load libraries and set options
#| warning: false
#| message: false
#| 

library(knitr)  
library(kableExtra)
library(tidyverse)


```

Read the data and take a first look

```{r}
#| label: read data - glimpse 

strawberry <- read_csv("strawberries25_v3.csv", col_names = TRUE)

glimpse(strawberry)
```

I have 12699 rows and 21 columns.

All I can see from the glimpse is I have date, location, values and coefficients of variation.

Examine the data. How is it organized?

```{r}
#| label: explore organization 1 


## Is every line associated with a state?

state_all <- strawberry |> distinct(State)

state_all1 <- strawberry |> group_by(State) |> count()

## every row is associated with a state

if(sum(state_all1$n) == dim(strawberry)[1]){print("Yes every row in the data is associated with a state.")}

## rm(state_all, state_all1)

##Ditch counties
unique(strawberry$`Geo Level`)

strawberry <- strawberry |> 
  filter(`Geo Level`== "NATIONAL" | `Geo Level`== "STATE")
  

```

## remove columns with a single value in all rows

```{r}
#|label: function def - drop 1-item columns

drop_one_value_col <- function(df){ 
 # browser()
  df_id <- ensym(df)
  msg = paste("Looking for single value columns in data frame: ",as.character(df_id) )
  print(msg)
  ## takes whole dataframe
dropc <- NULL
val <- NULL
## test each column for a single value
for(i in 1:dim(df)[2]){   
  if(dim(distinct(df[,i]))[1] == 1){
    dropc <- c(dropc, i)
    val <- c(val, df[1,i])
  }
} 

if(is.null(dropc)){
  print("No columns dropped")
  return(df)}else{
   print("Columns dropped:")
   # print(colnames(df)[drop])
   print(unlist(val))
   df <- df[, -1*dropc]
   return(df)
   }
}

## use the function

strawberry <- drop_one_value_col(strawberry)

drop_one_value_col(strawberry)

```

## Census data first

```{r}
#| label: srawberries split census, survey


straw_cen <- strawberry |> filter(Program=="CENSUS")

straw_sur <- strawberry |> filter(Program=="SURVEY")

straw_cen <- straw_cen |> drop_one_value_col()

straw_sur <- straw_sur |> drop_one_value_col()


```

```{r}
#| label: straw_cen split cols

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = " - ",
                         names = c("strawberries",
                                 "Category"),
                         too_many = "error",
                         too_few = "align_start"
                       )

```

```{r}
#| label: isolate organic

# 
# unique(straw_cen$strawberries)
# straw_cen$strawberries |> str_which("STRAWBERRIES") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC, FRESH MARKET") |> length()
# straw_cen$strawberries |> str_which("STRAWBERRIES, ORGANIC, PROCESSING" ) |> length()
# 
# ## count the cases
# straw_cen$strawberries |> str_which("ORGANIC") |> length()
# straw_cen$strawberries |> str_which("FRESH MARKET") |> length()
# straw_cen$strawberries |> str_which("PROCESSING") |> length()

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = strawberries,
                         delim = ", ",
                         names = c("strawberries",
                                 "ORGANIC",
                                 "organic_detail"),
                         
                         too_many = "error",
                         too_few = "align_start"
                       )


straw_cen <- straw_cen |> drop_one_value_col()

## how many organic rows?

organic_cen <- straw_cen |> filter(ORGANIC == "ORGANIC")

sum(is.na(straw_cen$ORGANIC))

#Subste straw_cen to non-organic only
straw_cen <- straw_cen[(is.na(straw_cen$ORGANIC)),]


straw_cen <- straw_cen |> drop_one_value_col()

```

Note that straw_cen has only one year: 2022

Current stats Census date has been isolated and split between Organic and Conventional strawberries

############################################ 

```{r}
#| label: explore straw_cen$Category

straw_cen <- straw_cen |>
  separate_wider_delim(  cols = `Category`,
                         delim = " ",
                         names = c("COL1",
                                 "COL2"),
                         too_many = "merge",
                         too_few = "align_start"
                       )


straw_cen$COL2 <- str_replace(straw_cen$COL2,"WITH ","")

straw_cen <- straw_cen |> rename(Measure = COL1, Bearing_type= COL2)

```
```{r}
#| label: explore straw_cen$Domain & Domain Category


## remove AREA GROWN and parens 
## change NOT SPECIFIEC TO TOTAL

straw_cen <- straw_cen |> rename(size_bracket = `Domain Category`)

straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "NOT SPECIFIED", "TOTAL")

straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "AREA GROWN: ", "")
straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "\\(", "")
straw_cen$size_bracket <- str_replace(straw_cen$size_bracket, "\\)", "")

```

```{r}
#| label: explore organic_cen

organic_cen <- organic_cen |> drop_one_value_col()

```
```{r}
#Split measure into acres and operations

Measure_acres_cen <- straw_cen |> filter(straw_cen$Measure == "ACRES")
Measure_acres_cen <- Measure_acres_cen |>
  separate_wider_delim(  cols = "size_bracket",
                         delim = " ",
                         names = c("Min Acres",
                                 "Max Acres"),
                         too_many = "merge",
                         too_few = "align_start"
                       )

Measure_acres_cen$`Max Acres` <- str_replace(Measure_acres_cen$`Max Acres`, "TO ", "")
Measure_acres_cen$`Max Acres` <- str_replace(Measure_acres_cen$`Max Acres`, "ACRES", "")
Measure_acres_cen$`Min Acres` <- str_replace(Measure_acres_cen$`Min Acres`, "TOTAL", "")

#Convert acres columns to double
Measure_acres_cen$`Max Acres` <- as.double(Measure_acres_cen$`Max Acres`)
Measure_acres_cen$`Min Acres` <- as.double(Measure_acres_cen$`Min Acres`)

#convert values column to numeric
Measure_acres_cen$Value <- str_replace(Measure_acres_cen$Value, ",","")
Measure_acres_cen$Value <- as.numeric(Measure_acres_cen$Value)

#convert CV% to CV and to double"
Measure_acres_cen$CV <- Measure_acres_cen$`CV (%)`
Measure_acres_cen <- Measure_acres_cen %>% select(-`CV (%)`)
Measure_acres_cen$CV <- as.double (Measure_acres_cen$CV)
Measure_acres_cen$CV <- (Measure_acres_cen$CV)/100

#sort acres min
Measure_acres_cen$`Min Acres`[is.na(Measure_acres_cen$`Min Acres`)] <- 0
Measure_acres_cen <- Measure_acres_cen[order(Measure_acres_cen$`Min Acres`),]

#divide into state and national 
state_cens <- Measure_acres_cen |> filter(Measure_acres_cen$`Geo Level`== "STATE")
state_cens$`State ANSI` <- as.numeric(state_cens$`State ANSI`)
state_cens <- state_cens[order(state_cens$`State ANSI`, state_cens$Bearing_type, state_cens$`Min Acres`),]

#impute state_cens NA values in col Value, using polynomial regression fit
impute_lm <- function(data){
  filter_data <- data %>% filter(`Min Acres`!=0, !is.na(`Min Acres`), !is.na(Value))
  uniq_acres <- unique(filter_data$`Min Acres`)
  if(length(uniq_acres)>1){
    model <- lm(Value ~ poly(`Min Acres`, 2, raw = TRUE), data = filter_data,
                na.action = na.exclude)
    data$Value[is.na(data$Value)] <- predict(model, newdata =
                                               data[is.na(data$Value),])
  }else{
    cat("Invalid")
  }
  
  return(data)
}

state_cens_imp <- state_cens %>% 
  group_by(`State ANSI`, Bearing_type) %>%
  do(impute_lm(.))

state_cens_imp$Value <- round(state_cens_imp$Value)


```



#############
To get better look at the data, look at California.

```{r}
#| label: explore California only

calif <- strawberry |>  filter(State=="CALIFORNIA")

## look at the unique values in the "Program" column

## in the consol
## unique(calif$Program)

## and look at the data selection widget on 
##      https://quickstats.nass.usda.gov

## You can see that CENSUS AND SURVEY are the two sources
## of data. (Why? What's the differences?).  So, let's see
## they differ.

calif_census <- calif |> filter(Program=="CENSUS")

calif_survey  <- calif |>  filter(Program=="SURVEY")

### do the same for Florida

FL <- strawberry |> filter(State == "FLORIDA")
FL_cens <- FL |> filter(Program == "CENSUS")
FL_surv <- FL |> filter(Program == "SURVEY")


##calif_survey <- strawberry |> select(Year, Period, `Data Item`, Value)


```

Explore California to understand the census and survey

```{r}
#| label: explore Calif census and survey 

## no assignment -- just exploring

calif_census_dropped <- drop_one_value_col(calif_census)

calif_survey_dropped <- drop_one_value_col(calif_survey)

#Explore Florida
FL_cens_dr <- drop_one_value_col(FL_cens)

FL_surv_dr <- drop_one_value_col(FL_surv)

```

Conclusions from California data exploration.

## Now return to the entire data set.

take the lessons learned by examining the California data

Two strategies -- columns first, rows first

Split the census data from the survey data. drop single value columns

```{r}
#Split census data from survey data of strawberry

strawberry_census <- strawberry |> filter(Program == "CENSUS")
strawberry_survey <- strawberry |> filter(Program == "SURVEY")

#drop single value columns

strawberry_census_dropped <- drop_one_value_col(strawberry_census)
strawberry_survey_dropped <- drop_one_value_col(strawberry_survey)

```

```{r}
strawberry_census$Value <- as.numeric(str_replace(strawberry_census$Value,",",""))
```

## separate composite columns

### `Data Item` into (fruit, category, item)

```{r}
#|label: split Data Item

  strawberry <- strawberry |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = ",",
                         names = c("Fruit",
                                 "Category",
                                 "Item",
                                 "Metric"),
                         too_many = "error",
                         too_few = "align_start"
                       )

strawberry_survey_split <- strawberry_survey_dropped |>
  separate_wider_delim( cols = `Data Item`,
                        delim = ",",
                        names = c("Fruit","Category","Item", "Metric"),
                        too_many = "error",
                        too_few = "align_start"
                       )

strawberry_survey_split$Category <- str_trim(strawberry_survey_split$Category, side = "both")
strawberry_survey_split$Item <- str_trim(strawberry_survey_split$Item, side = "both")
strawberry_survey_split$Metric <- str_trim(strawberry_survey_split$Metric, side = "both")

survey_chem <- strawberry_survey_split |>
  separate_wider_delim( cols = `Domain Category`,
                        delim = ",",
                        names = c("Chem", "Chem Usage"),
                        too_many = "error",
                        too_few = "align_start"
                       )  

survey_chem <- strawberry_survey_split|>
  separate_wider_delim(cols = `Chem Usage`,
                       delim = ":",
                       names = c("Chem Usage", "Chem Type"),
                       too_many = "error",
                      too_few = "align_start"
  )
## Use too_many and too_few to set up the separation operation.
# strawberry_census_split <- strawberry_census_dropped |>
#   separate_wider_delim( cols = `Data Item`,
#                         delim = ",",
#                         names = c("Category","Item", "Metric"),
#                         too_many = "debug",
#                         too_few = "align_start"
#                        )
# 
# strawberry_census_split$Category <- str_trim(strawberry_census_split$Category, side = "both")
# strawberry_census_split$Item <- str_trim(strawberry_census_split$Item, side = "both")
# strawberry_census_split$Metric <- str_trim(strawberry_census_split$Metric, side = "both")
# 

```

There is a problem you have to fix -- a leading space.

```{r}
#|label: fix the leading space

 # note
strawberry$Category[1]
# strawberry$Item[2]
# strawberry$Metric[6]
# strawberry$Domain[1]
## 
## trim white space 

strawberry$Category <- str_trim(strawberry$Category, side = "both")
strawberry$Item <- str_trim(strawberry$Item, side = "both")
strawberry$Metric <- str_trim(strawberry$Metric, side = "both")

```

## now exam the Fruit column -- find hidden sub-columns

```{r}

unique(strawberry_survey_split$Fruit)

## generate a list of rows with the production and price information

spr <- which((strawberry_survey_split$Fruit=="STRAWBERRIES - PRODUCTION") | (strawberry_survey_split$Fruit=="STRAWBERRIES - PRICE RECEIVED"))


strw_prod_price <- strawberry_survey_split |> slice(spr)

strw_chem <- strawberry_survey_split |> slice(-1*spr) 

  
```

## now examine the rest of the columns

Which ones need to be split?

## split sales and chemicals into two dataframes

(do this last after separating rows into separate data frames) (THEN rename the columns to correspond the analysis being done with the data frames)

```{r}
#|label: split srawberry into census and survey pieces

strw_b_sales <- strawberry |> filter(Program == "CENSUS")

strw_b_chem <- strawberry |> filter(Program == "SURVEY")

nrow(strawberry) == (nrow(strw_b_chem) + nrow(strw_b_sales))

## Move marketing-related rows in strw_b_chem 
## to strw_b_sales


```

## plots

```{r}
#|label: plot 1

plot1_data <- strawberry |> 
  select(c(Year, State, Category, Value)) |> 
  filter((Year == 2021) & (Category == "ORGANIC - OPERATIONS WITH SALES"))

plot1_data$Value <- as.numeric(plot1_data$Value)

plot1_data <- plot1_data |> arrange(desc(Value))

ggplot(plot1_data, aes(x=reorder(State, -Value), y=Value)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  labs(x = "States", y = "Count",
title ="Number of Organic Strawberry operations with Sales in 2021")


```

```{r}
## plot 2

plot2_data <- strawberry |> 
  select(c(Year, State, Category, Item, Value)) |> 
  filter((Year == 2021) & 
           (Category == "ORGANIC - SALES") & 
           (Item == "MEASURED IN $") & 
           (Value != "(D)"))


plot2_data$Value <- as.numeric(gsub(",", "", plot2_data$Value))

plot2_data <- plot1_data |> arrange(desc(Value))

ggplot(plot2_data, aes(x=reorder(State, -Value), y=Value)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  labs(x = "States", y = "Sales",
title ="Organic Strawberry Sales ($) in 2021")


```


#chem, fertilizer, production for survey
#for census: organic, nonorganic